---
title: "PA1_template.Rmd"
output: html_document
---

#Load necessary packages

```{r, echo = TRUE}
library(ggplot2)
library(dplyr)
library(lubridate)

```

#Explore the data

```{r}
url <-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
activity <- read.csv(unz("activity.zip", "activity.csv"), header=T, quote="\"", sep=",")

dim(activity)
str(activity)
names(activity)
summary(activity)
```
#Processing and Transforming the data

```{r}
nonaactivity <-na.omit(activity)
nonaactivity$date <- as.Date(nonaactivity$date)
str(nonaactivity)

```

#Feature Engineering - adding the day column, month and year columns

```{r}
nonaactivity <- transform(nonaactivity, day = day(nonaactivity$date))
nonaactivity <- transform(nonaactivity, month = month(nonaactivity$date))                                        
nonaactivity <- transform(nonaactivity, year = year(nonaactivity$date))
View(nonaactivity[1:500,])

```

#Provide the sum of total steps as grouped by date

```{r}
totals <- nonaactivity %>%
         group_by(date) %>%
         summarize(totalsteps = sum(steps))

```

#Create the histogram showing the total number of steps per day

```{r, echo=TRUE}
ggplot(nonaactivity, aes(date, steps)) + geom_bar(stat = "identity", colour = "black", fill = "black", width = 0.7) + facet_grid(. ~ month, scales = "free") + labs(title = "Total Number of Steps Each Day Per Month", x = "Date", y = "Total Steps")

```

#calculate the mean and the median

```{r}
mean(totals$totalsteps)
median(totals$totalsteps)
```

# What is the average daily activity pattern?
Calculate average steps for each of 5-minute interval during a 24-hour period.  
Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axi) and the average number of steps taken, averaged across all days (y-axis)  
Report which 5-minute interval, on average across all the days in the dataset contains the maximum number of steps?  
Observer and comment the average daily activity pattern

```{r}
library(plyr)
intervalsteps <- ddply(nonaactivity, ~ interval, summarize,mean=mean(steps))
qplot(x=interval, y=mean, data = intervalsteps,  geom = "line",
      xlab="Interval - 5 Minute",
      ylab="Step Count",
      main="Average Steps"
)
```

#Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
intervalsteps[which.max(intervalsteps$mean), ]
```

#Calculate and report the total number of missing values in the dataset

```{r}
activity_NA <- sum(is.na(activity))
activity_NA
```

#Imputing NAs
Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r}
newsteps <- aggregate(steps ~ interval, data = activity, FUN = mean)
imputdata <- numeric()
for (i in 1:nrow(activity)) {
  obs <- activity[i, ]
  if (is.na(obs$steps)) {
    steps <- subset(newsteps, interval == obs$interval)$steps
  } else {
    steps <- obs$steps
  }
  imputdata <- c(imputdata, steps)
}

imputactivity <- activity
imputactivity$steps <- imputdata

imputsteps <- aggregate(steps ~ date, data = imputactivity, sum, na.rm = TRUE)
```

```{r, echo=FALSE}
hist(imputsteps$steps, main = "Steps Per day", xlab = "Day")
```

```{r}
mean(imputsteps$steps)
median(imputsteps$steps)

```

#Are there differences in the weekdays vs weekends?

```{r}
weekenddata <- nonaactivity 
weekenddata$weekdays <- factor(format(weekenddata$date, "%A"))
levels(weekenddata$weekdays)
levels(weekenddata$weekdays) <- list(weekday = c("Monday", "Tuesday",
                                                 "Wednesday", 
                                                 "Thursday", "Friday"),
                                     weekend = c("Saturday", "Sunday"))
levels(weekenddata$weekdays)
table(weekenddata$weekdays)

stepaverage <- aggregate(weekenddata$steps, 
                         list(interval = as.numeric(as.character(weekenddata$interval)), 
                              weekdays = weekenddata$weekdays),
                         FUN = "mean")
names(stepaverage)[3] <- "meanOfSteps"
library(lattice)

```

```{r, echo=FALSE}
xyplot(stepaverage$meanOfSteps ~ stepaverage$interval | stepaverage$weekdays, 
       layout = c(1, 2), type = "l", 
       xlab = "Interval", ylab = "Number of steps")
```
